Place of death in NRS COVID-19 deaths: visualisation for 2022 data

### 2022-02-23 covid_nrs_place_of_death_visualisations_2022.Rmd
added weekly proportions of causes of death by place - interseting results w.r.t. cancer deaths in 2020 and 2021,with home deaths overtaking hospital deaths

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(ggtext)  # for showing coloured title text
```

@@

```{r}
figure_proportion_annual_deaths_by_place <-
  merged_proportions_of_deaths_by_place %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(size = 1) +
  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year))) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  # theme_minimal() +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(
    x = "Year", y = "Proportion of yearly deaths (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2022_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

@@

```{r}
# TODO: fix causes of deaths and reorder by number of deaths!
figure_cause_of_death_annual_proportion_by_place <-
  current_and_historical_deaths_by_cause_and_location %>%
  filter(!cause_of_death %in% c("All")) %>%
  filter(!year %in% c("2020-2021")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
  ) %>%
  ggplot(aes(x = year, y = proportion, colour = place_of_death, group = place_of_death)) +
  geom_line() +
  geom_point() +
  geom_text(data = ~filter(.x, year==2022), aes(x=4.2, y=proportion, label="*"), inherit.aes = FALSE) +  # 4.2 here is 4th (final) place on the scale!
  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE) +
  scale_colour_viridis_d() +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Year",
    y = "Percentage yearly deaths by place (%)",
    title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
    subtitle = paste0("\n* ",only_2022_text),
    caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",death_registrations_text,source_nrs_text), collapse = "\n")
  ) +
  theme(
    panel.spacing = unit(1, units = "lines"),
    legend.position = "top", legend.title = element_blank(),
    plot.title = element_textbox_simple()  # to render the html in the title
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

@@

>>>>

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(ggtext)  # for showing coloured title text

## load functions
source("./functions.R")
```

@@

```{r}
figure_proportion_annual_deaths_by_place <-
  merged_proportions_of_deaths_by_place %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(size = 1) +
  geom_text(data = ~filter(.x, year==2022), aes(x=2022.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year))) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  # theme_minimal() +
  theme(
    legend.position = "top", 
    plot.caption = element_text(size = 10, colour = "gray60")
    ) +
  labs(
    x = "Year", y = "Proportion of yearly deaths (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2022_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

@@

```{r}
figure_cause_of_death_annual_proportion_by_place <-
  current_and_historical_deaths_by_cause_and_location %>%
  filter(!cause_of_death %in% c("All")) %>%
  filter(!year %in% c("2020-2021")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
  ) %>%
  ggplot(aes(x = year, y = proportion, colour = place_of_death, group = place_of_death)) +
  geom_line() +
  geom_point() +
  geom_text(data = ~filter(.x, year==2022), aes(x=4.2, y=proportion, label="*"), inherit.aes = FALSE) +  # 4.2 here is 4th (final) place on the scale!
  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE) +
  scale_colour_viridis_d() +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Year",
    y = "Percentage yearly deaths by place (%)",
    title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
    subtitle = paste0("\n* ",only_2022_text),
    caption = paste0(c("The 2015-2019 number of deaths is the annual mean for the period.",death_registrations_text,source_nrs_text), collapse = "\n")
  ) +
  theme(
    panel.spacing = unit(1, units = "lines"),
    legend.position = "top", legend.title = element_blank(),
    plot.title = element_textbox_simple()  # to render the html in the title
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

@@

This is the same idea as above, but looking at weekly proportions of cause of death by place - where did deaths due to different causes happen, and did their proportion vary over the year (e.g. seasonal effects)

```{r}
figure_cause_of_death_weekly_proportion_by_place <-
  merged_deaths_cause_of_death %>%
  filter(!cause_of_death %in% c("All")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other", "All"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
  ) %>%
  group_by(year, week_number, cause_of_death, place_of_death != "All") %>%
  mutate(proportion_weekly_total = number_of_deaths / sum(number_of_deaths)) %>%
  ungroup %>%
  filter(place_of_death != "All") %>%
  ggplot(aes(x = date_w_c, y = proportion_weekly_total, colour = place_of_death, group = place_of_death)) +
  geom_line() +
  annotate(geom = "text", x=ymd("2021-01-14"), y=0.95, label = "2021", hjust = 0, colour = "grey40") +
  annotate(geom = "text", x=ymd("2022-01-14"), y=0.95, label = "'22", hjust = 0, colour = "grey40") +
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2021), linetype = "dashed", colour="grey50") +  # start of 2021
  geom_vline(xintercept = compute_start_date_from_week_number(week_number = 1, year_number = 2022), linetype = "dashed", colour="grey50") +  # start of 2022
  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE, ncol = 2) +
  scale_colour_viridis_d() +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2022)) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Year",
    y = "Percentage weekly deaths by place (%)",
    # title = "In 2020 and 2021 the share of deaths at <b style='color:#31688E;'>Home</b> increased for all causes.\nDuring the pandemic, more cancer deaths happened at <b style='color:#31688E;'>Home</b> than in <b style='color:#440154;'>Hospital</b>.",
    subtitle = paste0("\n* ",only_2022_text),
    caption = paste0(c("The 2015-2019 figures are average weekly deaths.",death_registrations_text,source_nrs_text), collapse = "\n")
  ) +
  theme(
    panel.spacing = unit(1, units = "lines"),
    legend.position = "top", legend.title = element_blank(),
    plot.title = element_textbox_simple(),  # to render the html in the title
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

figure_cause_of_death_weekly_proportion_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
```

### With 2015-2019 context also

```{r}
figure_weekly_proportion_of_death_by_place_with_2015_2019 <-
  bind_rows(
    ## 2020 onwards
    merged_deaths_cause_of_death %>%
      mutate(
        year = as.character(year),
        week_number_run_over = week_number_run_over + 53  # push 2020+ forward by a year, to 'squeeze' in the historic period year
        )
    ,
    ## 2015-2019 data, weekly averages make up one "year"
    merged_deaths_cause_of_death %>%
      filter(year==2020) %>%
      mutate(
        year = "2015-2019",
        number_of_deaths = deaths_mean
        )
  ) %>%
  mutate(year = factor(year, levels = c("2015-2019",as.character(2020:2022)))) %>%
  filter(!cause_of_death %in% c("All")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home", "Other", "All"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(
   cause_of_death = factor(cause_of_death %>% str_wrap(., width = 15), levels = order_of_causes_of_death %>% str_wrap(., width = 15)),  # needed to allow converting to a number for plotting the line segments; str_wrap breaks the text up into lines
  ) %>%
  group_by(year, week_number, cause_of_death, place_of_death != "All") %>%
  mutate(proportion_weekly_total = number_of_deaths / sum(number_of_deaths)) %>%
  ungroup %>%
  filter(place_of_death != "All") %>%
  ggplot(aes(x = week_number_run_over, y = proportion_weekly_total, colour = place_of_death, group = place_of_death)) +
  geom_line() +
  # annotate(geom = "text", x=1+1, y=0.95, label = "2015-19", hjust = 0, colour = "grey40") +
  # annotate(geom = "text", x=54+1, y=0.95, label = "2020", hjust = 0, colour = "grey40") +
  # annotate(geom = "text", x=107+1, y=0.95, label = "2021", hjust = 0, colour = "grey40") +
  # annotate(geom = "text", x=159+1, y=0.95, label = "'22", hjust = 0, colour = "grey40") +
  facet_rep_wrap(~cause_of_death, repeat.tick.labels = TRUE, ncol = 2) +
  scale_colour_viridis_d() +
  scale_x_continuous(breaks = c(1,54,107,159), labels = c("2015-2019",as.numeric(2020:2022))) +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Year",
    y = "Percentage weekly deaths by place (%)",
    subtitle = paste0("\n* ",only_2022_text),
    caption = paste0(c("The 2015-2019 figures are average weekly deaths",death_registrations_text,source_nrs_text), collapse = "\n")
  ) +
  theme(
    panel.spacing = unit(1, units = "lines"),
    legend.position = "top", legend.title = element_blank(),
    plot.title = element_textbox_simple(),  # to render the html in the title
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL

figure_weekly_proportion_of_death_by_place_with_2015_2019 %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_cause_of_death_weekly_proportion_by_place_2015_2019", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 25, height = 30, dpi = 300, units = "cm")
```
